---
title: "Basic usage of NIMBLE"
subtitle: "UW 2020 short course"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found 
library(nimble)
```

# Running example: An occupancy model

* AHM = [Applied Hierarchical Modeling in Ecology, Vol. I](https://www.mbr-pwrc.usgs.gov/pubanalysis/keryroylebook/) by Marc Kéry and J. Andrew Royle. 2015. Elsevier.
* Most AHM examples have been converted to NIMBLE: [https://github.com/nimble-dev/AHMnimble](https://github.com/nimble-dev/AHMnimble)
* Thanks to Marc Kéry, Andy Royle, and Mike Meredith for permission to post modified versions of their code on GitHub.
* Occupancy example from section 10.4:

    - Simulated data
    - `M` sites.
    - Each site is visited `J` times.
    - `y[i, j]` is detection (`1`) or non-detection (`0`) for visit `j` to site `i`.
    - Explanatory variables:

        - `vegHt` = Vegetation height: logistic effect on occupancy probability
        - `wind` = Wind speed: logistic effect on detection probability

# Occupancy example: Write the model code **in R**

- Slightly modified from AHM.
- Derived quantities are removed.

```{r echo = TRUE}
library(nimble)
occ_code <- nimbleCode({
  # Priors
  mean.p ~ dunif(0, 1)         # Detection intercept on prob. scale
  alpha0 <- logit(mean.p)      # Detection intercept
  alpha1 ~ dunif(-20, 20)      # Detection slope on wind
  mean.psi ~ dunif(0, 1)       # Occupancy intercept on prob. scale
  beta0 <- logit(mean.psi)     # Occupancy intercept
  beta1 ~ dunif(-20, 20)       # Occupancy slope on vegHt
  
  # Likelihood
  for (i in 1:M) {
    # True state model for the partially observed true state
    z[i] ~ dbern(psi[i])      # True occupancy z at site i
    logit(psi[i]) <- beta0 + beta1 * vegHt[i]
    for (j in 1:J) {
      # Observation model for the actual observations
      y[i,j] ~ dbern(p.eff[i,j])    # Detection-nondetection at i and j
      p.eff[i,j] <- z[i] * p[i,j]   # 'straw man' for WinBUGS
      logit(p[i,j]) <- alpha0 + alpha1 * wind[i,j]
    }
  }
  # Derived quantities are removed.
})
```

# Occupancy example: Simulate data

(This code is modified from AHM.)

```{r, fig.cap = '', fig.width=10, fig.height=6}
DO_PLOT <- TRUE
if(!exists("DO_PLOT"))
  DO_PLOT <- FALSE

# Choose sample sizes and prepare obs. data array y
set.seed(1)                   # So we all get same data set
M <- 100                      # Number of sites
J <- 3                        # Number of presence/absence measurements
y <- matrix(NA, nrow = M, ncol = J) # to contain the obs. data

# Create a covariate called vegHt
vegHt <- sort(runif(M, -1, 1)) # sort for graphical convenience

# Choose parameter values for occupancy model and compute occupancy
beta0 <- 0                    # Logit-scale intercept
beta1 <- 3                    # Logit-scale slope for vegHt
psi <- plogis(beta0 + beta1 * vegHt) # Occupancy probability
# plot(vegHt, psi, ylim = c(0,1), type = "l", lwd = 3) # Plot psi relationship

# Now visit each site and observe presence/absence perfectly
z <- rbinom(M, 1, psi)        # True presence/absence

# Look at data so far
table(z)

# Plot the true system state
if(DO_PLOT) {
  par(mfrow = c(1, 2), mar = c(5,5,2,2), cex.axis = 1.5, cex.lab = 1.5)
  plot(vegHt, z, xlab="Vegetation height", ylab="True presence/absence (z)",
              frame = F, cex = 1.5)
  plot(function(x) plogis(beta0 + beta1*x), -1, 1, add=T, lwd=3, col = "red")
}

# Create a covariate called wind
wind <- array(runif(M * J, -1, 1), dim = c(M, J))

# Choose parameter values for measurement error model and compute detectability
alpha0 <- -2                        # Logit-scale intercept
alpha1 <- -3                        # Logit-scale slope for wind
p <- plogis(alpha0 + alpha1 * wind) # Detection probability
# plot(p ~ wind, ylim = c(0,1))     # Look at relationship

# Take J = 3 presence/absence measurements at each site
for(j in 1:J) {
  y[,j] <- rbinom(M, z, p[,j])
}
sum(apply(y, 1, max))               # Number of sites with observed presences

# Plot observed data and true effect of wind on detection probability
if(DO_PLOT) {
  plot(wind, y, xlab="Wind", ylab="Observed det./nondetection data (y)",
             frame = F, cex = 1.5)
  plot(function(x) plogis(alpha0 + alpha1*x), -1, 1, add=T, lwd=3, col = "red")
}
# Look at the data: occupancy, true presence/absence (z), and measurements (y)
cbind(psi=round(psi,2), z=z, y1=y[,1], y2=y[,2], y3=y[,3])

# Bundle and summarize data set
str( occupancy_data <- list(y = y, 
                            vegHt = vegHt,
                            wind = wind,
                            M = nrow(y),
                            J = ncol(y),
                            XvegHt = seq(-1, 1, length.out=100),
                            Xwind = seq(-1, 1, length.out=100)) )

# Initial values: Best to give for same quantities as priors given.
zst <- apply(y, 1, max)        # Avoid data/model/inits conflict
occupancy_inits <- function(){
  list(z = zst, 
       mean.p = runif(1), 
       alpha1 = runif(1), 
       mean.psi = runif(1), 
       beta1 = runif(1))
}

# If running multiple chains, you'd use the function.
# We'll be running single chains for simplicity and wanting
# the same initial values for comparing different MCMC configurations.
set.seed(1)
inits_saved <- occupancy_inits()
```


# Occupancy Example: Steps to use NIMBLE:


1. Build the model.  It is an R object.
2. Build the MCMC.

    - 2a. Configure the MCMC.
    - 2b. Customize the MCMC.
    - 2c. Build the MCMC.

3. Compile the model and MCMC.
4. Run the MCMC (possibly multiple chains) and extract the samples.

- `nimbleMCMC` does all of this at once.
- Steps 2a-2c can be combined if no customization is needed.

    
# Occupancy Example: 1. Build the model


```{r}
occ_model <- nimbleModel(occ_code,
                        constants = occupancy_data, 
                        inits = occupancy_inits())
```

`constants` here are what JAGS calls `data`. We'll see in a bit that in NIMBLE we prefer to distinguish `data` from `constants`. 

# Occupancy Example: 2. Build the MCMC

```{r}
occ_MCMC <- buildMCMC(occ_model)
```

# Occupancy Example: 3. Compile the model and MCMC.

Compilation of the model and MCMC can be done together or separately. We'll do it separately.

```{r}
Cocc_model <- compileNimble(occ_model)
Cocc_MCMC <- compileNimble(occ_MCMC, project = occ_model)
```

# Occupancy Example: 4. Run the MCMC and get the samples

```{r}
samples <- runMCMC(Cocc_MCMC, niter = 2500)
```


# Occupancy Example: Look at results:


There are many packages for summarizing and plotting MCMC samples.  NIMBLE does not try to re-invent these wheels.

1. `mcmcplots`

```{r eval = FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "occupancy_samples_mcmcplot")
```

```{r echo = FALSE}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "orig_occupancy_samples_mcmcplot")
```

Results that comes with these slides are [here](orig_occupancy_samples_mcmcplot.html).

Results if you generated your own will be [here](occupancy_samples_mcmcplot.html).

2. `coda`

```{r eval = FALSE}
library(coda)
pdf("occupancy_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

```{r echo = FALSE}
library(coda)
pdf("orig_occupancy_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

Results that comes with these slides are [here](orig_occupancy_samples_coda.pdf).

Results if you generated your own will be [here](occupancy_samples_coda.pdf).

# MCMC workflow in NIMBLE

Here is a summary of the MCMC workflow in NIMBLE.

[nimbleMCMCworkflow.pdf](nimbleMCMCworkflow.pdf)

# Occupancy Example: Doing it all at once with `nimbleMCMC`


Start from:

- code
- constants + data
- inits

```{r}
samples <- nimbleMCMC(occ_code,
                       constants = occupancy_data,
                       inits = occupancy_inits,
                       niter = 10000,
                       nburnin = 1000,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
summary(samples) ## from coda
```

# Other notes


- `readBUGSmodel()` will read BUGS/JAGS model code and variables from their standard file formats.

